Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-10-24 paper/ms.tex

The functional form of $K_j$ may be used to approximate some commonly used Kernel components: radial
kernels can be approximated with $\beta_I=0$, while periodic kernels can be approximated by
$\beta_I \ne 0$.  The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be 
used to approximate each of these kernels, with various degrees of success:
\begin{itemize}
\item {\bf Constant kernel} A constant kernel component can be used to capture long-timescale
correlations in data, and can be represented with $\beta =  0$.
\item {\bf Exponential kernel.} This is also known as a Mat\'ern 1/2 kernel, and is exactly
represented with $\beta_I=0$.
\item The squared exponential kernel has the behavior of declining steeply at large
time separation, and has $c_1$ = 0.  An infinite number of exponential kernels may be used to
approximate a Gaussian kernel.  For example, to

\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ kernel is given by: 
\begin{equation}
K_{3/2}(\tau,\tau_0) = (1+x)e^{-x},
\end{equation} 
where $x = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier Transform\footnote{Using the normalization convention: 
$\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
\begin{equation}
\hat K_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
\end{equation}
where $\omega = 2\pi f$.  This kernel may be well approximated by the sum of two exponential kernels: 
\begin{equation}
K_{3/2}(\tau,\tau_0) \approx (1-b)e^{-x}+be^{-x\frac{b-1}{b}},
\end{equation} 
where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
large values of $b$, the maximum error on this scales as $2/(e^2b) \approx 0.27/b$; for
$b=10$, the error in this approximation is $<3$\%.  Note that the two exponential terms
have opposite signs, which may lead to numerical instability for values of $b$ that
are large.

One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
have this property, which leads to qualitatively different noise properties on small
timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
turns out to be an excellent description of stellar activity and granulation, and thus the approximate
formula is simultaneously physically relevant and computationally convenient.  
The power spectrum of a single exponential scales as $f^{-2}$ at high frequency,
while the Mat\'ern kernel scales as $f^{-4}$ which more closely matches the power
spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,2009A&A...495..979M}.
%\item Matern 5/2 kernel.
\item {\bf Cosine kernel.} The cosine kernel may be represented exactly with $\beta_R=0$.
\item exp-sin-square kernel
\end{itemize}

>>>>

The functional form of $K_j$ may be used to approximate some commonly used Kernel components: radial
kernels can be approximated with $\beta_I=0$, while periodic kernels can be approximated by
$\beta_I \ne 0$.  The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be 
used to approximate each of these kernels, with various degrees of success:
\begin{itemize}
\item {\bf Constant kernel} A constant kernel component can be used to capture long-timescale
correlations in data, and can be represented with $\beta =  0$.
\item {\bf Exponential kernel.} This is exactly
represented with $\beta_I=0$.
\item {\bf Cosine kernel.} The cosine kernel may be represented exactly with $\beta_R=0$.
\item exp-sin-square kernel
\item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $G(z)=e^{-z^2/2}$,
has the behavior of declining steeply at large time separation, is an even function, and has $c_1$ = 0.
Nevertheless, it may be well approximated by the sum of three exponential kernels:
\begin{eqnarray}
G(z) &\approx& 2.751807 e^{-1.531250 z} \cos{(0.341489 z)} \\
 &-& 0.354626 e^{-4.753504 z} \cos{(2.042572 z)}\\
 &-& 1.396834 e^{-1.796743 z} \cos{(1.764168 z)},
\end{eqnarray}
which has an accuracy of better than $<4\times 10^{-4}$ for all z.
%[2.751807496353032,1.5312499408182438,-0.34148888180984543,-0.35462563655835033,4.753504292338051,2.0425716395901157,-1.3968344706423566,1.7967431129750422,-1.7641684073374964
\item {\bf Mat\'ern $p+1/2$ kernel.}  The Mat\'ern kernel is given by:
\begin{equation}
C_{p+1/2}(x) = \sigma^2 e^{-\sqrt{(2p+1)}x} \frac{\Gamma(p+1)}{\Gamma(2p+1)}
\sum_{i=0}^p \frac{(p+i)!}{i!(p-i)!} (\sqrt{(8p+4)}x)^{p-i},
\end{equation} 
where $p$ is a positive integer.  In the limit $p\rightarrow \infty$, this becomes the
squared exponential kernel: $C_\infty(x) = \sigma^2 e^{-x^2/2}$.  But, for small
values of $p$, the Mat\'ern kernel may be approximated well by the sum of
$2p$ exponential kernels.  The exponential kernel is the Mat\'ern 1/2 kernel, while
$p=1$ is relevant to stellar variability, which we describe next.
\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ kernel is given by: 
\begin{equation}
K_{3/2}(\tau,\tau_0) = (1+x)e^{-x},
\end{equation} 
where $x = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier Transform\footnote{Using the normalization convention: 
$\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
\begin{equation}
\hat K_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
\end{equation}
where $\omega = 2\pi f$.  This kernel may be well approximated by the sum of two exponential kernels: 
\begin{equation}
K_{3/2}(\tau,\tau_0) \approx (1-b)e^{-x}+be^{-x\frac{b-1}{b}},
\end{equation} 
where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
large values of $b$, the maximum error on this scales as $2/(e^2b) \approx 0.27/b$; for
$b=10$, the error in this approximation is $<3$\%.  Note that the two exponential terms
have opposite signs, which may lead to numerical instability for values of $b$ that
are large.

One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
have this property, which leads to qualitatively different noise properties on small
timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
turns out to be similar to the observed properties of stellar activity and granulation, and thus the approximate
formula is simultaneously physically relevant and computationally convenient.  
The power spectrum of a single exponential scales as $f^{-2}$ at high frequency,
while the Mat\'ern kernel scales as $f^{-4}$ which more closely matches the power
spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,2009A&A...495..979M}.
\item {\bf Mat\'ern 5/2 kernel.}  The Mat\'ern 5/2 kernel is given by
\begin{equation}
C_{5/2}(y) = \sigma^2 (1+y+y^2/3)e^{-y},
\end{equation}
where $y = \sqrt{5}x$.  The Fourier transform is:
\begin{equation}
\hat K_{5/2}(\omega) = \frac{2\tau_0 10^{5/2}}{3\pi^{1/2}} (5+\tau_0^2\omega^2)^{-3}.
\end{equation}
This kernel may be approximated well by the sum of four exponential kernels
\begin{eqnarray}
K_{5/2}(y) &\approx& -147.6965573 e^{-1.0097610 y} + 85.8875034 e^{-1.0621734 y}\\
 &-&4.6540017 e^{-0.8353745 y} + 67.4630744 e^{-0.9160394 y},
\end{eqnarray}
which is accurate to $<4\times 10^{-6}$ for $y \le 10$.
The Fourier transform has a steeper dependence at large frequencies, $f^{-6}$, and as we are not aware of a stellar
power spectrum that requires a steeper spectrum than this, we forgo approximating further Mat\'ern kernels.
\end{itemize}

A standard practice in constructing kernels for solving problems with Gaussian Process
is to add or multiply the preceding standard kernel components.  In the case of damped cosine
kernels, the product of two kernels is: 
\begin{eqnarray}
K_j(\tau)K_k(\tau) &=&
\frac{1}{2} a_k a_j e^{-(\beta_{R,j}+\beta_{R,k}) \tau} \left(\cos{(\beta_{I,j}-\beta_{I,k})\tau}+\cos{(\beta_{I,j}+\beta_{I,k})\tau}\right)\\
&=& \frac{1}{2} \left(K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}-\beta_{I,k}) +K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}+\beta_{I,k})\right).
\end{eqnarray}
which is the sum of two damped cosine kernels.  Thus, only addition is necessary to
construct new kernels from damped cosine kernels.
